---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ausbushfire

`ausbushfire` is an R package specifically designed for analyzing the relationship between Australian bushfires and climate data. It provides a set of functions and an interactive Shiny application to help users easily explore global temperature change trends over the past century.

## Installation

You can install the development version of ausbushfire like so:

``` r
# Install remotes if needed
# install.packages("remotes")

remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-rabbit292929")
```

## Quick start

Launch the Shiny app bundled in the package:

```{r}
library(ausbushfire)
#run_ausbushfire_app()
```

Explore helper functions:

```{r}
# Descriptive statistics for 2000–2020
calculate_statistics("climate", 2000, 2020)
calculate_statistics("fwi",     2000, 2020)

# Annual trends (yearly mean / min / max)
annual_trends("climate", 2000, 2020)

# Extreme events per year (e.g., FWI ≥ 10)
extreme_events("fwi", threshold = 10, comparator = ">=", start_year = 2000, end_year = 2020)

```

## What’s included

### Shiny app

-   Fully interactive exploration of monthly / quarterly / yearly structure

-   Uses **package datasets only** (no `read.csv` inside the app)

-   Clear field descriptions and interpretation notes

-   Custom styling (Bootswatch + optional CSS)

-   Start with `run_ausbushfire_app()`

### Datasets

-   `climate_data` — monthly global **temperature anomalies** (°C), 1880–2025\

    Columns: `Year` (int), `Month` (abbrev `Jan`–`Dec`), `Temperature_Anomaly` (numeric), `Date` (mid-month `Date`), `Quarter` (`"Q1"`–`"Q4"`).

-   `fwi_data` — monthly **Fire Weather Index (FWI)** (synthetic/demo), 1900–2025\

    Columns: `Year`, `Month` (abbrev), `FWI`, `Date` (mid-month), `Quarter`.

<!-- -->

-   Columns: `Year`, `Month` (abbrev), `FWI`, `Date` (mid-month), `Quarter`.

> Note: `fwi_synthetic` is **synthetic demo data** for teaching and interactive examples, not directly computed from ERA5/CFFDRS.

### Helper functions

-   `calculate_statistics(data_type, start_year, end_year)` — descriptive stats within a year range

-   `annual_trends(data_type, start_year, end_year)` — yearly mean/min/max

-   `extreme_events(data_type, threshold, comparator, start_year, end_year)` — per-year extreme counts & frequency

## Documentation website (pkgdown)

**Website:** <https://ETC5523-2025.github.io/assignment-4-packages-and-shiny-apps-rabbit292929>

-   **Reference**: function documentation and dataset topics

-   **Articles**: “Getting started with ausbushfire” vignette

-   **Home**: brief description and quick links

## Reproducible data preparation

Raw data are cleaned under `data-raw/` and saved with `usethis::use_data()` into the package:

-   `data-raw/data-cleaning.R` reads the original CSV (e.g., `GLB.Ts+dSST.csv`), tidies monthly anomalies, derives SH seasonal summaries if needed, and saves the final objects.

-   All objects and functions are documented with **roxygen2**.

## License

This package is released under the **MIT License**. See [LICENSE](LICENSE).

## Acknowledgements

-   Temperature anomalies derived from NASA GISTEMP (e.g., `GLB.Ts+dSST.csv`), cleaned for teaching/demo.

-   FWI data in this package are **synthetic** and intended solely for instructional purposes.

## Maintainer

**ZHUO ZHANG** — [zzha0623\@student.monash.edu](mailto:zzha0623@student.monash.edu){.email}\

Issues & feature requests: \<<https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-rabbit292929/ausbushfire/issues>\>
